## Color Analysis Markdown
v15.07.26  
John Parkinson  
jparkinson@psu.edu

This R markdown file reproduces the analysis from:  
**"A citizen science approach to monitoring bleaching in the zoantharian 
*Palythoa tuberculosa*"**  
Parkinson JE, Yang SY, Kawamura I, Byron G, Todd P, Reimer JD 

*Please make sure to set the working director to the folder containing the
original data files:  
'parkinson_all_data.txt', 'hibino_bubble_data.txt', 
'hibino_color_data.txt', and 'hibino_nzpercent_data.txt* 

### Load libraries
```{r Load Libraries,echo=TRUE}
# setwd("PATH/TO/FOLDER/CONTAINING/DATA/FILES")
setwd("~/Desktop/final_color_analysis/") # Please change this path as needed

suppressMessages(library('dplyr')) # To tidy data
suppressMessages(library('reshape2')) # To reshape variable data
suppressMessages(library('psych')) # For creating correlation coefficient graphs
suppressMessages(library('irr')) # For the meanrho function
suppressMessages(library('bootstrap')) # To jackknife Krippendorff's alpha
suppressMessages(library('ggplot2')) # For graphs
suppressMessages(library('grid')) # For multiplot function
suppressMessages(library('gridExtra')) # For arranging Grobs to export
```

### Load data
```{r Load Data, echo=TRUE}

# All January 2014 data
data_all <- data.frame(read.table("parkinson_all_data.txt",head=TRUE))

# Hibino et al. color data
color_hibino <- data.frame(
        read.table("hibino_color_data.txt",head=TRUE,check.names=FALSE))

# Hibino et al. NZ percent data
nz_hibino <- data.frame(read.table("hibino_nzpercent_data.txt",
        head=TRUE,check.names=FALSE))

# Hibino et al. bubble plot data
data_bub <- read.table("hibino_bubble_data.txt",head=TRUE,check.names=FALSE)
```

### Test for observer reliability and correlation
```{r Test for observer reliability and correlation, echo=TRUE}

# Functions from Derek Jones:
# <http://www.r-bloggers.com/
#       agreement-between-code-readability-ratings-given-by-students/>

# Test observer reliability with Krippendorf's alpha

kripp.alpha_theta = function(x, xdata)
        {
        kripp.alpha(as.matrix(xdata[x,]), method="interval")$value
        }

jk_kripp.alpha = function(cs_subj)
        {
        cs_jackknife = jackknife(1:nrow(cs_subj), kripp.alpha_theta, cs_subj)
        quantile(cs_jackknife$jack.values, c(0.05, 0.95))
        }


# Data for testing observer reliability and correlation
data_obs <- select(data_all,obs0:obs11)

test_alpha <- kripp.alpha(t(data.matrix(data_obs)), method="interval")$value
 test_alpha 

test_alpha_jk <- jk_kripp.alpha(t(data.matrix(data_obs)))
 test_alpha_jk

#Test observer correlation with Spearman's rho

meanrho_theta = function(x, xdata)
        {
        meanrho(xdata[x,])$value
        }

jk_meanrho = function(subj)
        {
         cs_jackknife = jackknife(1:nrow(subj), meanrho_theta, subj)
         quantile(cs_jackknife$jack.values, c(0.05, 0.95))
        }

test_rho <- meanrho(data_obs)$value
 test_rho

test_rho_jk <- jk_meanrho(data_obs)
 test_rho_jk

```

### Test for correlations among variables
```{r Test for correlations among variables, echo=TRUE}

# Data for testing variable correlations
data_var <- select(data_all,colorscore,chlorophylla,symbiontdensity,nzpercent)

test_var <- corr.test(data_var, method="spearman", adjust="holm", alpha=0.05)
 test_var$r # Spearman's rho values
 test_var$p # Adjusted p-values (above the diagonal)

# To visualize quickly
# test_cor_plot <- pairs.panels(data_cor,
#       smooth=FALSE,scale=FALSE,ellipses=FALSE,lm=TRUE,method="spearman",
#       cor=TRUE,hist.col="white",rug=FALSE)
```

### Create plots
```{r Create plots, echo=TRUE}

# Color data
data_col <- select(data_all,colormean,colonylabel,colorcvmin,colorcvmax,
        colormin,colormax)
 data_col <- arrange(data_col,colormean)
 new_order <- data_col$colonylabel # Reorder for graphing
 data_col$colonylabel <- factor(data_col$colonylabel, levels=new_order)

# Chlorophyll a data
data_chl <- select(data_all,colorscore,chlorophylla)
 data_chl <- melt(data_chl, id.vars="colorscore",na.remove = FALSE)

# Density data
data_den <- select(data_all,colorscore,symbiontdensity)
 data_den <- melt(data_den, id.vars="colorscore", na.remove = FALSE)
 data_den <- mutate(data_den,value=value/100000) # Rescale for graphing

# NZ percent data
data_nzp <- select(data_all,colorscore,nzpercent)
 data_nzp <- melt(data_nzp, id.vars="colorscore", na.remove = FALSE)

theme_set(theme_bw(base_size = 20))  

plot_col <- ggplot(data=data_col, aes(x=colormean, y=colonylabel)) + 
        geom_point(aes(size=5)) +
        scale_x_continuous(limits = c(3.0, 6.0), breaks=seq(3.0, 6.0, 0.5)) +
        geom_vline(xintercept=c(3.25,3.75,4.25,4.75,5.25,5.75), 
                   linetype="dotted", color="grey") +
        geom_errorbarh(aes(xmin=colorcvmin, xmax=colorcvmax), size=1, height=0.5) +
        geom_errorbarh(aes(xmin=colormin, xmax=colormax), linetype=2, height=0) +
        labs(x="Color Score\n", y="colony", title="A.") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_nzp <- ggplot(data=data_nzp, aes(x=colorscore, y=value)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(colorscore)), binaxis="y", 
                stackdir="center", binwidth=0.8) +
        scale_fill_manual(values = c("black","black","black","black","black")) +
        scale_x_continuous(limits = c(3.0, 6.0), breaks=seq(3.0, 6.0, 0.5)) +
        scale_y_continuous(limits = c(70, 90), breaks=seq(70, 90, 5)) +
        geom_vline(xintercept=c(3.25,3.75,4.25,4.75,5.25,5.75), 
                   linetype="dotted", color="grey") +
        annotate("text", x=6.0, y=90,hjust=1,vjust=1,
                 label="rho = -0.027\np-adj = 1.00") +
        labs(x="Color Score\n", y="normal zooxanthellae (%)", title="B.") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_den <- ggplot(data=data_den, aes(x=colorscore, y=value)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(colorscore)), binaxis="y", 
                stackdir="center", binwidth=0.8) +
        scale_fill_manual(values = c("black","black","black","black","black")) +
        scale_x_continuous(limits = c(3.0, 6.0), breaks=seq(3.0, 6.0, 0.5)) +
        scale_y_continuous(limits = c(0, 25), breaks=seq(0, 25, 5)) +
        geom_vline(xintercept=c(3.25,3.75,4.25,4.75,5.25,5.75), 
                   linetype="dotted", color="grey") +
        annotate("text", x=6.0, y=25,hjust=1,vjust=1,
                 label="rho = -0.030\np-adj = 1.00") +
        labs(x="Color Score\n", 
             y=bquote('cells' ~cm^-3~ '('~x10^6*')'),title="C.") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_chl <- ggplot(data=data_chl, aes(x=colorscore, y=value)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(colorscore)), binaxis="y", 
                stackdir="center", binwidth=0.8) +
        scale_fill_manual(values = c("black","black","black","black","black")) +
        scale_x_continuous(limits = c(3.0, 6.0), breaks=seq(3.0, 6.0, 0.5)) +
        scale_y_continuous(limits = c(0, 25), breaks=seq(0, 25, 5)) +
        geom_vline(xintercept=c(3.25,3.75,4.25,4.75,5.25,5.75), 
                 linetype="dotted", color="grey") +
        annotate("text", x=6.0, y=25,hjust=1,vjust=1,
                 label="rho = -0.472\np-adj = 0.215") +
        labs(x="Color Score\n", 
             y=bquote('chlorophyll'~italic(a)~mu*g~cm^-3),title="D.") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )
```

## Define and execute multiplot function
```{r Define and execute multiplot function, echo=TRUE, fig.width=12, fig.height=10}

# Multiple plot function from: 
# <http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/>

multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  require(grid)

  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  if (is.null(layout)) {
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    for (i in 1:numPlots) {
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

multiplot(plot_col,plot_den,plot_nzp,plot_chl,cols=2)

# To export as .png for publication
# Fig2 <- arrangeGrob(plot_col,plot_nzp,plot_den,plot_chl)
# ggsave(filename="Figure2_final.png", 
#       plot=Fig2, scale=2,width=6.6, height=5.3, units="in")
```

## Calculate Hibino correlations
```{r Calculate Hibino correlations, echo=TRUE, fig.width=12, fig.height=10}

# The bubble plot below shows the distribution of color scores over 2009

data_bub <- read.table("hibino_bubble_data.txt",head=TRUE,check.names=FALSE)
 data_bub <- melt(data=data_bub,
                id="Month",variable.name="Score",value.name="Freq.")
 data_bub <- subset(data_bub,Freq. != 0) # Remove zero values
 data_bub$Month <- with(data_bub,
                factor(Month,levels=rev(levels(Month)))) # Reverse months
      
plot_bub <- ggplot(data_bub,aes(Score, Month, size=Freq.)) + 
        geom_point(shape=21, fill="black") +
        scale_size_continuous(range=c(3,16)) +
        scale_y_discrete(labels=c('Nov','Oct','Sep','Aug','Jul','Jun',
                'May','Apr','Feb','Jan')) +
        labs(x="Color Score\n", y="", title="A. Color Score Distribution 2009") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "left"
        )

# This calculates monthly correlations.

data_monthly <- cor(color_hibino[,2:13],nz_hibino[,2:13], 
                       method = "spearman") # Only interested in the diagonal
 monthly_rho <- diag(data_monthly)
 monthly_rho 

monthly_p <- rep(NA,13)
        for (i in 2:13) {
        temp <- cor.test(color_hibino[,i],nz_hibino[,i])
        monthly_p[i] <- temp$p.value
        }
 monthly_p <- monthly_p[-1]
 monthly_p

# Based on this, April, May, June, and July are significant (all of these points have 
# correlations > 0.6, so by adding a horizontal line at 0.6, we separate the 
# significant from the non-significant correlations)

# This plot shows correlations with NZ% in a time series

monthly_pair <- cbind(c(10:1),monthly_rho[c(1:2,4:11)])
 colnames(monthly_pair) <- c("month","rho")
 monthly_pair <- data.frame(monthly_pair)
 
plot_monthly <- ggplot(data=monthly_pair, aes(x=rho, y=month)) + 
        geom_point(aes(size=5)) +
        scale_size_continuous(range=c(3,12)) +
        scale_x_continuous(limits = c(-0.2, 1), breaks=seq(-0.2, 1, 0.4)) +
        scale_y_continuous(limits = c(1,10), breaks=seq(1,10,1), 
           labels=c('Nov','Oct','Sep','Aug','Jul','Jun','May','Apr','Feb','Jan')) +
        geom_vline(xintercept=0.6,linetype="dashed", color="black") +
        labs(x="Spearman's Rho\n", y="", title="B. Monthly NZ% Correlations 2009") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

multiplot(plot_bub,plot_monthly,cols=2)

# To export as .png for publication
# Fig3 <- arrangeGrob(plot_bub,plot_monthly,ncol=2)
# ggsave(filename="Figure3_final.png", 
#       plot=Fig3, scale=2,width=6.6, height=5.3, units="in")
```

## Make plots comparing to Hibino data
```{r Make plots comparing to Hibino data, echo=TRUE, fig.width=12, fig.height=10}

data_may09 <- data.frame(cbind(color_hibino[,6],nz_hibino[,6]))
colnames(data_may09)=c("color","nz")

data_jun09 <- data.frame(cbind(color_hibino[,7],nz_hibino[,7]))
colnames(data_jun09)=c("color","nz")

data_jan09 <- data.frame(cbind(color_hibino[,2],nz_hibino[,2]))
colnames(data_jan09)=c("color","nz")

# Note the January 2014 data is the same as in the first graph above.

plot_may09 <- ggplot(data=data_may09, aes(x=color, y=nz)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(color)), binaxis="y", 
                stackdir="center", binwidth=3) +
        scale_fill_manual(values=
                c("black","black","black","black","black","black")) +
        scale_x_continuous(limits = c(2.5, 5.5), breaks=seq(2.5, 5.5, 0.5)) +
        scale_y_continuous(limits = c(20, 120), breaks=seq(20, 100, 20)) +
        geom_vline(xintercept=c(2.75,3.25,3.75,4.25,4.75,5.25), 
                   linetype="dotted", color="grey") +
        annotate("text", x=5.5, y=20,hjust=1,vjust=0,
                 label="rho = 0.753\np-adj = 0.004") +
        labs(x="Color Score\n", y="normal zooxanthellae (%)", title="A. May 2009") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_jun09 <- ggplot(data=data_jun09, aes(x=color, y=nz)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(color)), binaxis="y", 
                stackdir="center", binwidth=3) +
        scale_fill_manual(values=
                c("black","black","black","black","black","black")) +
        scale_x_continuous(limits = c(2.5, 5.5), breaks=seq(2.5, 5.5, 0.5)) +
        scale_y_continuous(limits = c(20, 120), breaks=seq(20, 100, 20)) +
        geom_vline(xintercept=c(2.75,3.25,3.75,4.25,4.75,5.25), 
                   linetype="dotted", color="grey") +
        annotate("text", x=5.5, y=20,hjust=1,vjust=0,
                 label="rho = 0.726\np-adj = 0.009") +
        labs(x="Color Score\n", y="normal zooxanthellae (%)", 
             title="B. June 2009") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_jan09 <- ggplot(data=data_jan09, aes(x=color, y=nz)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(color)), binaxis="y", 
                stackdir="center", binwidth=3) +
        scale_fill_manual(values=
                c("black","black","black","black","black","black")) +
        scale_x_continuous(limits = c(2.5, 5.5), breaks=seq(2.5, 5.5, 0.5)) +
        scale_y_continuous(limits = c(20, 120), breaks=seq(20, 100, 20)) +
        geom_vline(xintercept=c(2.75,3.25,3.75,4.25,4.75,5.25), 
                   linetype="dotted", color="grey") +
        annotate("text", x=5.5, y=20,hjust=1,vjust=0,
                 label="rho = 0.347\np-adj = 0.515") +
        labs(x="Color Score\n", y="normal zooxanthellae (%)", 
             title="C. January 2009") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

plot_jan14 <- ggplot(data=data_nzp, aes(x=colorscore, y=value)) + 
        geom_smooth(method="lm",color="black",linetype="longdash") +
        geom_dotplot(aes(fill=as.factor(colorscore)), binaxis="y", 
                stackdir="center", binwidth=3) +
        scale_fill_manual(values=
                c("black","black","black","black","black","black")) +
        scale_x_continuous(limits = c(2.5, 5.5), breaks=seq(2.5, 5.5, 0.5)) +
        scale_y_continuous(limits = c(20, 120), breaks=seq(20, 100, 20)) +
        geom_vline(xintercept=c(2.75,3.25,3.75,4.25,4.75,5.25), 
                   linetype="dotted", color="grey") +
        annotate("text", x=5.5, y=20,hjust=1,vjust=0,
                 label="rho = -0.027\np-adj = 1.00") +
        labs(x="Color Score\n", 
             y="normal zooxanthellae (%)", title="D. January 2014") +
        theme(plot.title = element_text(size=20, face="bold", vjust=2, hjust=0)) +
        theme(
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.7),
                legend.position = "none",
                panel.grid.major = element_line(color="white"),
                panel.grid.minor = element_line(color="white"),
                panel.border = element_rect(color="black")
          )

multiplot(plot_may09,plot_jan09,plot_jun09,plot_jan14,cols=2)

# To export as .png for publication
# Fig4 <- arrangeGrob(plot_may09,plot_jun09,plot_jan09,plot_jan14)
# ggsave(filename="Figure4_final.png", 
#       plot=Fig4, scale=2,width=6.6, height=5.3, units="in")
```